---
title: NFL Big Data Bowl 2021 setup and play animation
author: Mike Morris
date: '2020-10-24'
slug: nfl-big-data-bowl-2021-exploration.en-us
categories:
  - NFL
tags:
  - NFL
  - R
  - Seahawks
keywords:
  - tech

---


The [2021 edition of the NFL's Big Data Bowl](https://www.kaggle.com/c/nfl-big-data-bowl-2021/overview) is here. The challenge is to evaluate defensive performance on passing plays using player tracking data, without offensive or defensive lines.  

Resources that I've found helpful :

* [Tom Bliss tutorial](https://www.kaggle.com/tombliss/tutorial)
* [Definitions of events](https://twitter.com/asmae_toumi/status/1317644112435216385) in the data
* How to [animate a specific play](https://github.com/nfl-football-ops/Big-Data-Bowl/blob/master/README.md)
* [R for NFL analytics](http://rstudio-pubs-static.s3.amazonaws.com/494188_154898728be3411ebab39050ca8a8dcd.html)



What's covered here:

* SQLite database of the tracking data
* examples of animations using the best and worst (by EPA) Seahawks plays in the dataset. 


The tracking data takes up a fair amount of space. Originally I started to import and merge the various csvs, but that crashed my R session repeatedly, so I decided to use a sqlite database where I can pull into memory only what I need. Here is the code that I used to create it, which is spread across 17 different csv files, along with files for players, games, and plays. 


```{r create sqlite db, eval=FALSE}


library(RSQLite)
library(dplyr)
library(dbplyr)
library(DBI)
library(tidyverse)


###create a database
trackingdata <- RSQLite::dbConnect(SQLite(), "trackingdata.sqlite")


###add tables from csv

#games file
games <- read_csv("2021/nfl-big-data-bowl-2021/games.csv")
dbWriteTable(trackingdata, "games", games)
rm(games)

#players file
players <- read_csv("2021/nfl-big-data-bowl-2021/players.csv")
dbWriteTable(trackingdata, "players", players)
rm(players)

#plays file 
#includes targeted receiver data not included in original data release
#targeted receiver data available here: https://www.kaggle.com/tombliss/nfl-big-data-bowl-2021-bonus
plays <- read_csv("2021/nfl-big-data-bowl-2021/plays.csv")
targeted_receivers <- read_csv("2021/nfl-big-data-bowl-2021/targetedReciever.csv")
plays <- merge(plays, targeted_receivers, by = c("gameId", "playId"))

dbWriteTable(trackingdata, "plays", plays)
rm(plays)
rm(targeted_receivers)


#check that we are doing ok so far
#list tables in a database
dbListTables(trackingdata)
# List columns in a table
dbListFields(trackingdata, "plays")


# tracking data, create table with week 1 data
tracking <- read_csv("2021/nfl-big-data-bowl-2021/week1.csv")


#Standardizing tracking data so its always in direction of offense vs raw on-field coordinates.
tracking <- tracking %>%
  mutate(x = ifelse(playDirection == "left", 120-x, x),
         y = ifelse(playDirection == "left", 160/3 - y, y))
dbWriteTable(trackingdata, "tracking", tracking)
rm(tracking)


#now do the same thing for weeks 2-17
for (w in 2:17){
  #temperory dataframe used for reading week for given iteration
  tracking <- read_csv(paste0("/home/michael/Documents/NFL/NFL big data bowl/2021/nfl-big-data-bowl-2021/week",w,".csv"),
                               col_types = cols())
  
  #Standardizing tracking data so its always in direction of offense vs raw on-field coordinates.
  tracking <- tracking %>%
    mutate(x = ifelse(playDirection == "left", 120-x, x),
           y = ifelse(playDirection == "left", 160/3 - y, y))
  
  #append to the sqlite db
  dbWriteTable(conn=trackingdata, name="tracking", tracking, append=T, row.names=F)
  rm(tracking)
}


```




I then made this function to animate any play in the database (see links above for the main code used to build this, I didn't create most of it):

```{r play animation function, echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE}

library(tidyverse)
library(gganimate)
library(cowplot)

  library(RSQLite)
  library(dplyr)
  library(dbplyr)
  library(DBI)
  library(tidyverse)


animate_passing_play <- function(gameID, playID){
  
  #open the connection
  trackingdata <- RSQLite::dbConnect(SQLite(), "/home/michael/Documents/NFL/NFL big data bowl/trackingdata.sqlite")
  
  query_tracking_result <- tbl(trackingdata, "tracking")
  query_play_result <- tbl(trackingdata, "plays")
  
  
  #get the data for only the play we are interested in
  single_play <- query_tracking_result %>% 
    dplyr::filter(playId == playID) %>% 
    dplyr::filter(gameId == gameID)  %>%
    collect()
  
  
  #get information about the play
  single_play_info <- query_play_result %>% 
    dplyr::filter(playId == playID) %>% 
    dplyr::filter(gameId == gameID)  %>%
    collect()
  
  plot_title <- str_trim(gsub("\\s*\\([^\\)]+\\)","",as.character(single_play_info$playDescription[1])))
  
  
  #disconnect the connection
  RSQLite::dbDisconnect(trackingdata)
  
  
  
  
  ## General field boundaries
  xmin <- 0
  xmax <- 160/3
  hash.right <- 38.35
  hash.left <- 12
  hash.width <- 3.3
  
  
  ## Specific boundaries for a given play
  ymin <- max(round(min(single_play$x, na.rm = TRUE) - 10, -1), 0)
  ymax <- min(round(max(single_play$x, na.rm = TRUE) + 10, -1), 120)
  df_hash <- expand.grid(x = c(0, 23.36667, 29.96667, xmax), y = (10:110))
  df_hash <- df_hash %>% filter(!(floor(y %% 5) == 0))
  df_hash <- df_hash %>% filter(y < ymax, y > ymin)
  
  
  
  animate_play <- ggplot() +
    scale_size_manual(values = c(6, 4, 6), guide = FALSE) + 
    scale_shape_manual(values = c(21, 16, 21), guide = FALSE) +
    scale_fill_manual(values = c("#e31837", "#654321", "#002244"), guide = FALSE) + 
    scale_colour_manual(values = c("black", "#654321", "#c60c30"), guide = FALSE) + 
    annotate("text", x = df_hash$x[df_hash$x < 55/2], 
             y = df_hash$y[df_hash$x < 55/2], label = "_", hjust = 0, vjust = -0.2) + 
    annotate("text", x = df_hash$x[df_hash$x > 55/2], 
             y = df_hash$y[df_hash$x > 55/2], label = "_", hjust = 1, vjust = -0.2) + 
    annotate("segment", x = xmin, 
             y = seq(max(10, ymin), min(ymax, 110), by = 5), 
             xend =  xmax, 
             yend = seq(max(10, ymin), min(ymax, 110), by = 5)) + 
    annotate("text", x = rep(hash.left, 11), y = seq(10, 110, by = 10), 
             label = c("G   ", seq(10, 50, by = 10), rev(seq(10, 40, by = 10)), "   G"), 
             angle = 270, size = 4) + 
    annotate("text", x = rep((xmax - hash.left), 11), y = seq(10, 110, by = 10), 
             label = c("   G", seq(10, 50, by = 10), rev(seq(10, 40, by = 10)), "G   "), 
             angle = 90, size = 4) + 
    annotate("segment", x = c(xmin, xmin, xmax, xmax), 
             y = c(ymin, ymax, ymax, ymin), 
             xend = c(xmin, xmax, xmax, xmin), 
             yend = c(ymax, ymax, ymin, ymin), colour = "black") + 
    geom_point(data = single_play, aes(x = (xmax-y), y = x, shape = team,
                                       fill = team, group = nflId, size = team, colour = team), alpha = 0.7) + 
    geom_text(data = single_play, aes(x = (xmax-y), y = x, label = jerseyNumber), colour = "white", 
              vjust = 0.36, size = 3.5) + 
    ylim(ymin, ymax) + 
    coord_fixed() +  
    theme_nothing() + 
    theme(plot.title = element_text()) +
    labs(title = plot_title) +
    transition_time(frameId)  +
    ease_aes('linear') + 
    NULL
  
  
  # animate_play
  
  
  ## Ensure timing of play matches 10 frames-per-second
  play.length.ex <- length(unique(single_play$frameId))
  animate(animate_play, fps = 10, nframe = play.length.ex)
  
  
}






```



Here are some examples. 

First, the Seahawks (on offense) play in the database with the best expected points added, a 50+ yard touchdown to Tyler Lockett.

```{r best Seahawks play by epa, echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}


#best SEA EPA pass play
sea_best_epa <- animate_passing_play(2018092312, 1886)
sea_best_epa

```


Second, the Seahawks offensive play with the worst expected points added, a pick six from midfield.


```{r worst Seahawks play by epa, echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE}
#worst SEA EPA pass play
sea_worst_epa <-animate_passing_play(2018091700, 3637)

sea_worst_epa

```




This is super cool stuff to play around with, I hope to do more in the coming weeks.

