---
title: Draft capital spent by position (2010-2019)
author: Mike Morris
date: '2020-01-17'
slug: draft-capital-by-position
categories:
  - NFL
tags:
  - NFL
keywords:
  - tech
---

<!--more-->

The purpose of this post is to examine:

* How much draft capital has been invested in different position groups over the last 10 years?
* Are there trends in investment?

Sources:

* [Draft history](https://www.pro-football-reference.com/draft/)
* [Draft pick values](https://www.footballperspective.com/draft-value-chart/)

Here I looked at results for the 2010-2019 drafts, covering the last 10 years. To value draft capital, I focused on the Chase Stuart model, but I report the Jimmy Johnson model results below as well.

This is subject to some uncertainty in terms of position. For instance, is a LB really an edge rusher? I am trusting positions from Pro Football Reference here.

```{r data, echo=FALSE, warning=FALSE, message = FALSE}


#load libraries
library(nflscrapR)
library(tidyverse)
library(dplyr)
library(na.tools)
library(rlang)
library(kableExtra)
library(rvest)

#read in draft history results by position back to 2009

#draft history

url <- "https://www.pro-football-reference.com/years/2019/draft.htm"

draft_2019 <- url %>%
  xml2::read_html() %>%
  html_nodes(xpath='//*[@id="drafts"]') %>%
  html_table()
draft_2019 <- draft_2019[[1]]

draft_2019 <- draft_2019[1:5] 
colnames(draft_2019) <- c("Round", "Pick", "Team", "Player", "Position")
draft_2019$Year <- 2019 
draft_2019 <- draft_2019 %>%
  dplyr::filter(Round !="Rnd")

#loop through other years, clean it up, and merge with 2019 file
for (y in 2010:2018){
  
  url <- paste0("https://www.pro-football-reference.com/years/", y, "/draft.htm")
  
  draft_results <- url %>%
    xml2::read_html() %>%
    html_nodes(xpath='//*[@id="drafts"]') %>%
    html_table()
  draft_results <- draft_results[[1]]
  
  draft_results <- draft_results[1:5] 
  colnames(draft_results) <- c("Round", "Pick", "Team", "Player", "Position")
  draft_results$Year <- y 
  draft_results <- draft_results %>%
    dplyr::filter(Round !="Rnd")
  
  
  draft_2019 <- rbind(draft_2019, draft_results)
  
}

draft_results <- draft_2019
rm(draft_2019)


#get the draft capital table

#draft values
url <- "https://www.footballperspective.com/draft-value-chart/"
draft_values <- url %>%
  xml2::read_html() %>%
  html_nodes(xpath='//*[@id="tablepress-276"]') %>%
  html_table()
draft_values <- draft_values[[1]]
colnames(draft_values) <- c("Pick", "ChaseStuartValue", "JimmyJohnsonValue")

draft_history_value <- merge(draft_results, draft_values, by = "Pick")



#compare to next years EPA/play
#load the data that I previously stored so I can work with it easily
# load("/home/michael/Documents/pbp_2010_to_2019.Rda")

#look at capital by position groups, offense total, defense total, special teams
draft_history_value <- draft_history_value %>%
  dplyr::mutate(GeneralPosition = dplyr::case_when(
    Position == "WR" ~ "WR",
    Position == "DE" ~ "DL",
    Position == "DB" ~ "DB",
    Position == "T" ~ "OL",
    Position == "DT" ~ "DL",
    Position == "LB" ~ "LB",
    Position == "QB" ~ "QB",
    Position == "RB" ~ "RB",
    Position == "CB" ~ "DB",
    Position == "G" ~ "OL",
    Position == "TE" ~ "TE",
    Position == "S" ~ "DB",
    Position == "OLB" ~ "LB",
    Position == "C" ~ "OL",
    Position == "ILB" ~ "LB",
    Position == "OL" ~ "OL",
    Position == "NT" ~ "DL",
    Position == "P" ~ "Specialist",
    Position == "K" ~ "Specialist",
    Position == "LS" ~ "Specialist",
    Position == "P" ~ "Specialist",
    Position == "DL" ~ "DL",
    Position == "FB" ~ "RB")) %>%
  dplyr::mutate(SideofBall = dplyr::case_when(
    GeneralPosition == "WR" ~ "Offense",
    GeneralPosition == "QB" ~ "Offense",
    GeneralPosition == "RB" ~ "Offense",
    GeneralPosition == "OL" ~ "Offense",
    GeneralPosition == "TE" ~ "Offense",
    GeneralPosition == "QB" ~ "Offense",
    GeneralPosition == "RB" ~ "Offense",
    GeneralPosition == "DL" ~ "Defense",
    GeneralPosition == "DB" ~ "Defense",
    GeneralPosition == "LB" ~ "Defense"))

```

Generally speaking teams have invested slightly more in defense than offense (NAs here are for specialists).

<br>

```{r investmentBySideofBall, echo=FALSE, warning=FALSE, message = FALSE}

#total captial invested by position group

total_by_sideofball <- draft_history_value %>%
  dplyr::group_by(SideofBall) %>%
  dplyr::summarise(sum(ChaseStuartValue), sum(JimmyJohnsonValue)) %>%
  dplyr::arrange(-`sum(ChaseStuartValue)`)

kableExtra::kable(total_by_sideofball,
                   caption = "NFL Draft Capital Invested by Side of Ball (2010-2019)")
```


<br>

By position group, teams have invested more in defensive positions, particularly the line and defensive backs.

<br>

```{r investmentByPositionGroup, echo=FALSE, warning=FALSE, message = FALSE}

#total captial invested by position group

total_by_position_group <- draft_history_value %>%
  dplyr::group_by(GeneralPosition) %>%
  dplyr::summarise(sum(ChaseStuartValue), sum(JimmyJohnsonValue)) %>%
  dplyr::arrange(-`sum(ChaseStuartValue)`)

kableExtra::kable(total_by_position_group,
                   caption = "NFL Draft Capital Invested by Position Group (2010-2019)")


```


<br>

Next I looked for linear trends in draft capital expended (correlations of totals for each group by year). For instance, are teams using more capital for offense, or defense, or certain positions over the past 10 years? For instance, are teams drafting fewer early running backs?

Let's take a look at offense vs defense first.

<br>

```{r trendsbyoffdef, echo=FALSE, warning=FALSE, message = FALSE}

position_value_by_year_offdef <- draft_history_value %>%
  dplyr::filter(SideofBall == "Offense" | SideofBall == "Defense" ) %>%
  dplyr::group_by(SideofBall, Year) %>%
  dplyr::summarise(sum(ChaseStuartValue), sum(JimmyJohnsonValue))

cor_position_val_by_year_offdef <- position_value_by_year_offdef %>%
  dplyr::group_by(SideofBall) %>%
  dplyr::summarise(cor(`sum(ChaseStuartValue)`, Year), cor(`sum(JimmyJohnsonValue)`, Year)) 
colnames(cor_position_val_by_year_offdef) <- c("SideofBall", "r_ChaseStuartValue_Year",  "r_JimmyJohnsonValue_Year")

cor_position_val_by_year_offdef <- cor_position_val_by_year_offdef %>%
  dplyr::arrange(-r_ChaseStuartValue_Year)



#general rounding function for all numeric valus in a dataframe
round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
  df[,nums] <- round(df[,nums], digits = digits)
  (df)
}

#round the correlations
cor_position_val_by_year_offdef <- round_df(cor_position_val_by_year_offdef, 2)



kableExtra::kable(cor_position_val_by_year_offdef,
                   caption = "NFL Draft Capital Trends by Side of Ball (2010-2019)")



```

<br>

No trends at all. Now let's look by position group:

<br>

```{r trendsbyposition, echo=FALSE, warning=FALSE, message = FALSE}

position_value_by_year <- draft_history_value %>%
  dplyr::group_by(GeneralPosition, Year) %>%
  dplyr::summarise(sum(ChaseStuartValue), sum(JimmyJohnsonValue))

cor_position_val_by_year <- position_value_by_year %>%
  dplyr::group_by(GeneralPosition) %>%
  dplyr::summarise(cor(`sum(ChaseStuartValue)`, Year), cor(`sum(JimmyJohnsonValue)`, Year)) 
colnames(cor_position_val_by_year) <- c("GeneralPosition", "r_ChaseStuartValue_Year","r_JimmyJohnsonValue_Year" )

cor_position_val_by_year <- cor_position_val_by_year %>%
  dplyr::arrange(-r_ChaseStuartValue_Year)



#general rounding function for all numeric valus in a dataframe
round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
  df[,nums] <- round(df[,nums], digits = digits)
  (df)
}

#round the correlations
cor_position_val_by_year <- round_df(cor_position_val_by_year, 2)



kableExtra::kable(cor_position_val_by_year,
                   caption = "NFL Draft Capital Trends by Position Group (2010-2019)")



```

<br>

I have to admit these results surprised me. If we are to interpret these it appears over this time period teams are spending more draft capital on linebackers and tight ends, and less of defensive line, but these trends are not statistically significant with N = 10. Running backs, if anything, appear to be just as likely to be picked early now as they were 10 years ago. And teams are spending just about the same amount on quarterback, despite widespread recognition that having a decent quarterback on a rookie contract is a huge organizational advantage.

In short, it is not at all clear that the passing revolution has impacted how teams approach drafting position groups.

